---
title: "Scientific journal subscription costs in Finland 2010-2015: a preliminary analysis"
author: "Leo Lahti"
date: "2016-06-10 18:00:00"
excerpt: "Freedom of Information"
layout: post
draft: false
categories: R
---


```{r foi-init, message=FALSE, echo=FALSE}
library(dplyr)
library(tidyr)
library(reshape2)
library(ggplot2)
theme_set(theme_bw(20))
# License fee data URL
# f <- "~/data/Tietopyynto/20160607\ Kustantajahintatiedot_koottu_v0.99.csv"
f = "http://avaa.tdata.fi/documents/kuhiti/Publisher_Costs_FI_Full_Data.csv"
# Read the data
df <- read.csv(f, fileEncoding = "UTF-8")
# Rename
df = rename(df, Publisher = Publisher.Supplier)
df = rename(df, Organization = Organization.name)
```

[Detailed information on journal subscription costs paid to individual
publishers by the Finnish research
institutions](http://openscience.fi/-/transparency-and-openness-to-scientific-publishing-the-finnish-research-organisations-pay-millions-of-euros-annually-to-the-large-publishers) has been released by the Finnish Ministry of Education and Culture, and its Open Science and Research Initiative funded 2014–2017 ([Kustantajahintatiedot Suomessa 2010–2015](http://urn.fi/urn:nbn:fi:csc-kata20160609091336769027)). 

With this, **Finland becomes to our knowledge
the first country where annual subscription fees for all individual
publishers and all major research institutions have been made
available**, spanning the years 2010-2015. Similar information has
been previously released for some, but not all publishers and research
institutions in the UK and US; and related activities are ongoing in
several countries (see the recent [blog post by Stuart
Lawson](http://stuartlawson.org/2016/06/publicly-available-data-on-international-journal-subscription-costs)).

Access to literature is fundamental to academic research but this has been challenged with the rapidly increasing prices of academic journals, which university libraries find increasingly difficult to fund. According to the [data from the US Association of Research Libraries](http://www.lib.washington.edu/scholpub/facts/economics) the academic journal subscription charges increased 4x faster than inflation in 1986-2007. The relative variation in prices is also considerable among the publishers  ([Bergstrom et al. PNAS 2014](http://www.econ.ucsb.edu/~tedb/Journals/PNAS-2014-Bergstrom-1403006111.pdf)). Limited access to detailed pricing information and agreement details are likely to result in suboptimal contracts ([Cockerill, 2006](https://olh.openlibhums.org/articles/10.16995/olh.72/#B7); [Shieber, 2009](https://olh.openlibhums.org/articles/10.16995/olh.72/#B66)). Improved access to subscription costs can hence be expected to lead to better deals and lower costs for the universities. It can also facilitate transition to the Open Access (OA) publishing model.

Motivated by all this, and following [Tim Gowers successful FOI request on
the subscription costs for scientific journals in
UK](http://gowers.wordpress.com/2014/04/24/elsevier-journals-some-facts/),
we made a similar request in Finland together with the [Open Knowledge
Finland association and other Open Science
advocates](https://www.facebook.com/groups/241398182642057/permalink/411482855633588). After
the Finnish universities [turned down my FOI request in summer
2014](https://github.com/okffi-science/2014-tietopyynto-lisenssimaksut),
we appealed in court, which decided the case positive for us in August
2015. For an English summary of this process, see a separate post in the [MostlyPhysics](http://www.mostlyphysics.net/blog/) blog.

Here I provide a brief preliminary analysis of the
[data](http://avointiede.fi/ajankohtaista/-/asset_publisher/UJglmibGKmbR/content/lapinakyvyytta-ja-avoimuutta-tieteelliseen-julkaisemiseen-tutkimusorganisaatioilta-vuosittain-miljoonia-euroja-suurille-kustantajille?_101_INSTANCE_UJglmibGKmbR_viewMode=view)
on journal subscription fees that was collected and released by the
Ministry of Education Open Science Initiative in Finland. I have
abbreviated some terms as detailed in the source code of
this analysis is [maintained in
Github](https://github.com/rOpenGov/2016-Finland-SubscriptionCosts).



## Overall subscription costs 2010-2015

Finland paid in total `r round(sum(df$Cost)/1e6, 1)` million EUR
subscription and other fees on scientific publishing in 2010-2015. The overall breakup of the costs is available as a separate [table](http://data.okf.fi/ropengov/20160613-FOI/dashboard.html). The
average annual costs for in Finland were `r round(mean((df %>%
group_by(Year) %>% summarise(Costs = sum(Cost)/1e6))$Costs))` MEUR in
2010-2015; this is one third of the annual subscription costs in
Austria (70 MEUR; [Bauer et al.,
2015](https://olh.openlibhums.org/articles/10.16995/olh.72/#B4)), and
two thirds of the annual expenditure (31 MEUR) in [New
Zealand](http://publicaddress.net/9549). Data for the top-10
publishers in the UK 2010-2014 is available in [Lawson, Meghreblian &
Brook,
2015](https://olh.openlibhums.org/articles/10.16995/olh.72/#B45)
([Table
1](https://olh.openlibhums.org/articles/10.16995/olh.72)). During this
period the UK paid altogether 4319 MEUR (rough estimate based on the
exchange rate June 12, 2016) for the top-10 publishers. Finland paid
61 MEUR for the same top-10 publishers in the same period, which is
roughly 17% of the UK expenditure per capita (unexpectedly low?). The costs
in the other countries seem unexpectedly high compared to Finland, I
wonder what is the explanation. Either Finland can negotiate
considerably better deals, or, perhaps more likely, the figures from
the different countries are not directly comparable but this will
require further investigation.

The data set covers all Finnish universities, major public
institutions, and a number of special libraries, information services
and other smaller institutions. Open access article processing charges
(APCs) are not included in this data collection as far as I can see,
although it would be interesting in its own right. It was recently
reported by [Naukkarinen
(2016)](https://avointiede.fi/documents/10864/12232/Avoimen+julkaisemisen+tuen+malli/73838e9b-7924-446c-9c7a-cc8f759919bb)
that in 2014 [18% of the articles in Finnish universities were
published as open
access](https://l.facebook.com/l.php?u=https%3A%2F%2Favointiede.fi%2Fdocuments%2F10864%2F12232%2FAvoimen%2Bjulkaisemisen%2Btuen%2Bmalli%2F73838e9b-7924-446c-9c7a-cc8f759919bb&h=-AQFlKomT). It
was also estimated that publishing all articles as Open Access would
have cost 17 MEUR, whereas the subscription fees in 2014 were 22
MEUR. This suggests that the transition to the Open Access model might
be a good idea. Overall, there are roughly 35 000 peer-reviewed
academic journals globally ([Ware & Mabe,
2015](https://olh.openlibhums.org/articles/10.16995/olh.72/#B71));
less than a third of these (11 000) are open access ([DOAJ,
2016](https://olh.openlibhums.org/articles/10.16995/olh.72/#B12)). A
common complaint regarding the (golden) OA model is the relatively
high cost of the article processing charges, typically paid from
primary research funding. With this funding model, money going OA is
frequently seen as being taken away from other research activities,
such as experiments or research personnell. In contrast, publication
costs in the conventional subscription model are funded through
university libraries with secret discloure agreements with the
publishers and hence largely masked from the research community. In
order to assess the true costs of scientific publishing and in order
to facilitate shift to OA, the costs of the subscription model must be
made more transparent and money used for conventional subscription
must be diverted to funding the costs of OA publishing.

The full data set details the subscription fees also by agreement
type, organization type, and [subscription
category](http://data.okf.fi/ropengov/20160613-FOI/dashboard.html).


## Costs by publisher

The Finnish data covers `r length(unique(df$Publisher))` individual publishers ([see annual costs by publisher](http://data.okf.fi/ropengov/20160613-FOI/dashboard.html)). The figure indicates the total subscription fees paid to the top publishers 2010-2015. Over one third of the total costs go to Elsevier, which has been often [criticized](https://gowers.wordpress.com/2014/04/24/elsevier-journals-some-facts/) for its huge [profit margins](http://journals.plos.org/plosone/article?id=10.1371%2Fjournal.pone.0127502). The costs are given per bundle, so we cannot compare individual publishers on a per article or per citation basis based on this data set. It was recently estimated elsewhere, however, that Elsevier’s prices per citation are roughly 3x higher than with non-profit publishers; Emerald, Sage, and Taylor & Francis had roughly 10x higher prices ([Bergstrom et al. PNAS 2014](http://www.econ.ucsb.edu/~tedb/Journals/PNAS-2014-Bergstrom-1403006111.pdf)).

```{r foi-totalcosts2b, message=FALSE, echo=FALSE, fig.width=10, fig.height=9}
dfs = df %>% group_by(Publisher) %>% summarise(Costs = sum(Cost)/1e6) %>% arrange(desc(Costs))
dfs$Publisher = factor(dfs$Publisher, levels = rev(as.character(dfs$Publisher)))
dfs$Percentage = round(100 * dfs$Costs/sum(dfs$Costs), 1)
p <- ggplot(dfs[1:20,], aes(x = Publisher, y = Costs)) + geom_bar(stat = "identity", fill = "lightgray", color = "black") + ggtitle("Costs by publisher") + coord_flip() + xlab("") + ylab("Total Costs (MEUR)") + geom_text(aes(x=Publisher, y=32, label=paste(Percentage, "%", sep = ""), hjust=ifelse(sign(Percentage)>0, 0, 0)), position = position_dodge(width=1), size = 7)
print(p)

dfs = df %>% group_by(Publisher) %>% summarise(n = n()) %>% arrange(desc(n))
dfs$Publisher = factor(dfs$Publisher, levels = rev(as.character(dfs$Publisher)))
dfs$Percentage = round(100 * dfs$n/sum(dfs$n), 1)
p <- ggplot(dfs[1:20,], aes(x = Publisher, y = n)) + geom_bar(stat = "identity", fill = "lightgray", color = "black") + ggtitle("Subscriptions by publisher") + coord_flip() + xlab("") + ylab("Subscriptions (n)") + geom_text(aes(x=Publisher, y=32, label=paste(Percentage, "%", sep = ""), hjust=ifelse(sign(Percentage)>0, 0, 0)), position = position_dodge(width=1), size = 7)
#print(p)

dfs = df %>% group_by(Resource.type) %>% summarise(Costs = sum(Cost)/1e6) %>% arrange(desc(Costs))
dfs$Resource.type = factor(dfs$Resource.type, levels = rev(as.character(dfs$Resource.type)))
dfs$Percentage = round(100 * dfs$Costs/sum(dfs$Costs), 1)
```


```{r foi-costperyear, message=FALSE, eval=TRUE, echo=FALSE, fig.show="hold", out.width="200px"}
dfs = df %>% group_by(Publisher) %>% summarise(Costs = sum(Cost)/1e6) %>% arrange(desc(Costs))
```

The total costs paid to scientific publishers by Finland have increased roughly 10% per year in 2010-2015 (annual increase is indicated in the left figure). The top-10 publishers correspond to `r round(100 * sum(dfs[["Costs"]][1:10])/sum(dfs[["Costs"]]))`% of the overall costs (right figure). See a separate table for full [annual costs by publisher](http://data.okf.fi/ropengov/20160613-FOI/dashboard.html) (the top-10 publishers shown below). 

```{r foi-costbytime, message=FALSE, echo=FALSE, fig.width=16, fig.height=6, out.width="870px"}
# Cost, Year : all publishers
# Growth percentage compared to previous year is shown
dfs = df %>% group_by(Year) %>% summarise(Costs = sum(Cost)/1e6) 
p1 <- ggplot(dfs, aes(x = Year, y = Costs)) +
       geom_bar(stat = "identity", fill = "lightgray", color = "black") + ggtitle("Total costs (Finland)") +
       geom_text(data = dfs[-1,], aes(x = Year, y = 1.5,
     label = paste(round(100 * diff(dfs$Costs)/dfs$Costs[-length(dfs$Costs)], 1), "%", sep = "")   ),
     	   size = 6) +
       scale_x_continuous(breaks = 2010:2015, labels = c(2010:2015)) + 
       ylab("Costs (MEUR)")
#print(p1)

dfss = df %>% group_by(Publisher) %>% summarise(Costs = sum(Cost, na.rm = TRUE)) %>% arrange(desc(Costs))
top.publishers <- as.character(dfss$Publisher[1:10])
dfs2 <- dplyr::filter(df, Publisher %in% top.publishers) %>% group_by(Year, Publisher) %>% summarise(Costs = sum(Cost)/1e6) %>% arrange(Year)
dfs2$Publisher <- factor(dfs2$Publisher, levels = top.publishers)
p2 <- ggplot(dfs2, aes(x = Year, y = Costs, color = Publisher)) +
       geom_point() +
       geom_line() +       
       ggtitle("Cost development by publisher") +
       ylab("Costs (MEUR)") 
#print(p2)

library(gridExtra)
grid.arrange(p1, p2, widths=c(5, 9))


dfs <- df %>% group_by(Year, Publisher) %>% summarise(Costs = sum(Cost, na.rm = TRUE)/1e6) %>% arrange(Year) %>% spread(Year, Costs)
dfs[is.na(dfs)] = 0
dfs$Publisher = as.character(dfs$Publisher)
dfs$Total = rowSums(dfs[, as.character(2010:2015)], na.rm = TRUE)
dfs$Percentage = 100 * dfs$Total/sum(dfs$Total)
dfs = dfs %>% arrange(desc(Percentage))
dfs = rbind(c("Total", colSums(dfs[, -1])), dfs)
names(dfs) = c("Publisher (costs in MEUR)", as.character(2010:2015), "Total", "%")
for (i in 2:ncol(dfs)) {dfs[,i] = round(as.numeric(unlist(dfs[,i])),2)}
kable(head(dfs, 10))
```  


```{r foi-timebypublisher2, message=FALSE, echo=FALSE}
# Normalize the costs by 2010 to highlight relative increases
dfs <- df %>% group_by(Year, Publisher) %>% summarise(Costs = sum(Cost, na.rm = TRUE)/1e6) %>% arrange(Year)
dfs = dfs %>% spread(Year, Costs)
dfs[, as.character(2010:2015)] = apply(as.matrix(dfs[, as.character(2010:2015)]), 2, function (x) {x/unlist(dfs[, "2010"])})
dfs = melt(dfs)
names(dfs) = c("Publisher", "Year", "Costs")
dfs$Year = as.numeric(as.character(dfs$Year))
```

Let us compare the [relative increase in publisher costs](http://data.okf.fi/ropengov/20160613-FOI/dashboard.html). The costs are normalized to 1 in 2010, and the top 10 publishers with the highest cost increase in 2010-2015 are shown. The `r sum(!(!is.na(subset(dfs, Year == 2010)$Costs) & !is.na(subset(dfs, Year == 2015)$Costs)))` publishers that did not have declared costs in 2010 or 2015 (see [here](http://data.okf.fi/ropengov/20160613-FOI/dashboard.html)) are excluded.

```{r foi-timebypublisher2b, message=FALSE, echo=FALSE, fig.width=14, fig.heigth=3}
# Just show publisher with the largest cost increases
dfs$Costs[dfs$Costs == 0] = NA
top <- as.character((subset(dfs, Year == 2015) %>% arrange(desc(Costs)))$Publisher)[1:10]
dfs <- dfs %>% dplyr::filter(Publisher %in% top)
# Order the factor levels for the figure
dfs$Publisher <- factor(dfs$Publisher, levels = rev(as.character((subset(dfs, Year == 2015) %>% arrange(Costs))$Publisher)))
p <- ggplot(dfs, aes(x = Year, y = Costs, color = Publisher)) +
geom_point() +
       geom_line() +       
       ggtitle("Relative cost increase by publisher") +
       ylab("Relative cost increase") +
       scale_y_continuous(breaks = seq(1, max(dfs$Costs), 2))
print(p)
```


<p></p>

## Costs by organization

```{r foi-totalcosts2c, message=FALSE, echo=FALSE}
dfs = df %>% group_by(Organization.type) %>% summarise(Costs = 100*sum(Cost)/sum(df$Cost))
```

The Finnish data collection includes `r length(unique(df$Organization))` organizations ([see annual costs by organization](http://data.okf.fi/ropengov/20160613-FOI/dashboard.html)). The universities ('yliopisto') are responsible of `r as.numeric(round(dfs[dfs$Organization.type == "University", "Costs"],1))`% of all costs (left figure); University of Helsinki had the highest total costs in 2010-2015 (`r round(as.numeric((dfs = df %>% group_by(Organization) %>% summarise(Costs = sum(Cost)/1e6) %>% arrange(desc(Costs)))[1, "Costs"]),1)` MEUR; top institutions shown in the right figure). The Table indicates the annual costs for the top organizations.

```{r foi-totalcosts2d, message=FALSE, echo=FALSE, fig.show="hold", out.width="420px", fig.width = 10, fig.height = 9}
dfs = df %>% group_by(Organization.type) %>% summarise(Costs = sum(Cost)/1e6) %>% arrange(Costs)
dfs$Organization.type = factor(dfs$Organization.type, levels = as.character(dfs$Organization.type))
dfs$Percentage = round(100 * dfs$Costs/sum(dfs$Costs), 1)
p <- ggplot(dfs, aes(x = Organization.type, y = Costs)) + geom_bar(stat = "identity", fill = "lightgray", color = "black") + ggtitle("Costs by organization type") + coord_flip() + xlab("") + ylab("Total Costs (MEUR)") + geom_text(aes(x=Organization.type, y=80, label=paste(Percentage, "%", sep = ""), hjust=ifelse(sign(Percentage)>0, 0, 0)), position = position_dodge(width=1), size = 7)
print(p)

dfs = df %>% group_by(Organization) %>% summarise(Costs = sum(Cost)/1e6) %>% arrange(desc(Costs))
dfs$Organization = factor(dfs$Organization, levels = rev(as.character(dfs$Organization)))
dfs$Percentage = round(100 * dfs$Costs/sum(dfs$Costs), 1)
p <- ggplot(dfs[1:20,], aes(x = Organization, y = Costs)) + geom_bar(stat = "identity", fill = "lightgray", color = "black") + ggtitle("Costs by organization") + coord_flip() + xlab("") + ylab("Total Costs (MEUR)") + geom_text(aes(x=Organization, y=20, label=paste(Percentage, "%", sep = ""), hjust=ifelse(sign(Percentage)>0, 0, 0)), position = position_dodge(width=1), size = 7)
print(p)


dfs <- df %>% group_by(Year, Organization) %>% summarise(Costs = sum(Cost, na.rm = TRUE)/1e6) %>% arrange(Year) %>% spread(Year, Costs)
dfs[is.na(dfs)] = 0
dfs$Organization = as.character(dfs$Organization)
dfs$Total = rowSums(dfs[, as.character(2010:2015)], na.rm = TRUE)
dfs$Percentage = 100 * dfs$Total/sum(dfs$Total)
dfs = dfs %>% arrange(desc(Percentage))
dfs = rbind(c("Total", colSums(dfs[, -1])), dfs)
names(dfs) = c("Organization (costs in MEUR)", as.character(2010:2015), "Total", "%")
for (i in 2:ncol(dfs)) {dfs[,i] = round(as.numeric(unlist(dfs[,i])),2)}
kable(head(dfs, 10))
```

```{r foi-timebyorganization2, message=FALSE, echo=FALSE, fig.width=14, fig.heigth=3}
# Normalize the costs by 2010 to highlight relative increases
dfs <- df %>% group_by(Year, Organization) %>% summarise(Costs = sum(Cost, na.rm = TRUE)/1e6) %>% arrange(Year)
dfs = dfs %>% spread(Year, Costs)
dfs[, as.character(2010:2015)] = apply(as.matrix(dfs[, as.character(2010:2015)]), 2, function (x) {x/unlist(dfs[, "2010"])})
dfs = melt(dfs)
names(dfs) = c("Organization", "Year", "Costs")
dfs$Year = as.numeric(as.character(dfs$Year))
```


Finally, let us compare the [relative increase in costs across the institutions](http://data.okf.fi/ropengov/20160613-FOI/dashboard.html). The costs are normalized to 1 in 2010, and the top 10 organizations with the highest cost increase in 2010-2015 are shown. The `r sum(!(!is.na(subset(dfs, Year == 2010)$Costs) & !is.na(subset(dfs, Year == 2015)$Costs)))` organizations that did not have declared costs in 2010 or 2015 (see [here](http://data.okf.fi/ropengov/20160613-FOI/dashboard.html)) are excluded.

```{r foi-timebyorganization2b, message=FALSE, echo=FALSE, fig.width=14, fig.heigth=3}
# Just show organization with the largest cost increases
dfs$Costs[dfs$Costs == 0] = NA
top <- as.character((subset(dfs, Year == 2015) %>% arrange(desc(Costs)))$Organization)[1:10]
dfs <- dfs %>% dplyr::filter(Organization %in% top)
# Order the factor levels for the figure
dfs$Organization <- factor(dfs$Organization, levels = rev(as.character((subset(dfs, Year == 2015) %>% arrange(Costs))$Organization)))
p <- ggplot(dfs, aes(x = Year, y = Costs, color = Organization)) +
geom_point() +
       geom_line() +       
       ggtitle("Cost development by organization") +
       ylab("Relative cost increase") +
       scale_y_continuous(breaks = seq(1, max(dfs$Costs), .25))
print(p)
```


