---
title: helsinki vignette
layout: tutorial_page
package_name: helsinki
package_name_show: helsinki
author: Leo Lahti, Juuso Parkkinen, Joona Lehtomaki
meta_description: Helsinki R Tools
github_user: ropengov
package_version: 0.9.09
header_descripton: Helsinki R Tools
---



Helsinki  R tools
===========

The helsinki R package is part of the
[rOpenGov](http://ropengov.github.com/helsinki) project.


### Installation

Release version for general users (NOT AVAILABLE YET!)

```{r install, eval=FALSE}
# install.packages("helsinki")
# library(helsinki)
```

Development version for developers:

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("helsinki", "ropengov")
library(helsinki)
```

Further installation and development instructions at the [home
page](http://ropengov.github.com/helsinki).


## Helsinki region environmental services (Helsingin seudun ympäristöpalvelut HSY)



### Population grid

Download population information from [HSY database](http://www.hsy.fi/seututieto/kaupunki/paikkatiedot/Sivut/Avoindata.aspx) ((C) 2011 HSY) and visualize on the Helsinki map. For data documentation, see [HSY website](http://www.hsy.fi/seututieto/Documents/Paikkatiedot/Tietokuvaukset_kaikki.pdf). Population size, density, and age structure per grid.

```{r hsy-example, fig.width=8, fig.height=8, message=FALSE}
# http://ropengov.github.com/helsinki
library(helsinki)

# Get population grid (Vaestotietoruudukko) 
sp <- get_HSY_data("Vaestotietoruudukko")  

# Visualize population grid
#library(gisfi)
# Define limits of the color scale
#at <- c(seq(0, 2000, 250), Inf) 
#q <- PlotShape(sp, "ASUKKAITA", type = "oneway", at = at, ncol = length(at)) 
```

Inspect data manually. Some rarely populated grids are censored with
'99' to guarantee privacy.

```{r hsy-example2, message=FALSE}
df <- as.data.frame(sp)   
head(df)
```

### Helsinki building information

Information of buildings in Helsinki region. Data obtained from (C)
HSY 2011. Grid-level (500mx500m) information on building counts
(lukumaara), built area (kerrosala), usage (kayttotarkoitus), region
efficiency (aluetehokkuus).

```{r hsy-examplerak, message=FALSE}
sp <- get_HSY_data("Rakennustietoruudukko")  
df <- as.data.frame(sp)   
head(df)
```

### Helsinki building area capacity

Building area capacity per municipal region (kaupunginosittain summattu tieto rakennusmaavarannosta). Data obtained from (C) HSY 2011. 

```{r hsy-exampleramava, message=FALSE}
sp <- get_HSY_data("seutuRAMAVA_kosa")  
df <- as.data.frame(sp)   
head(df)
```


### Further HSY data

[HSY website](http://www.hsy.fi/seututieto/kaupunki/paikkatiedot/Sivut/Avoindata.aspx) has data for 2010-2011. More is coming, add later and allow temporal analysis.


## Helsinki Real Estate Department (Helsingin kaupungin kiinteistövirasto, HKK)

Retrieve [HKK](http://kartta.hel.fi/avoindata/index.html) data sets.

### Helsinki address information

```{r hkk-addressH, message=FALSE}
dat <- get_HKK_address_data("Helsingin osoiteluettelo")
head(dat)
```

```{r hkk-addressS, message=FALSE}
dat <- get_HKK_address_data("Seudullinen osoiteluettelo")
head(dat)
```


### Licensing and Citations

This work can be freely used, modified and distributed under the
[Two-clause FreeBSD
license](http://en.wikipedia.org/wiki/BSD\_licenses). Cite Helsinki R
package and and the appropriate data provider, including a url
link. Kindly cite the R package as 'Leo Lahti, Juuso Parkkinen ja
Joona Lehtomäki (2013-2014). helsinki R package. URL:
http://ropengov.github.io/helsinki'.

For further usage examples, see
[Louhos-blog](http://louhos.wordpress.com) and
[takomo](https://github.com/louhos/takomo/tree/master/Helsinki).


### Session info


This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```

