---
title: helsinki vignette
layout: tutorial_page
package_name: helsinki
package_name_show: helsinki
author: Juuso Parkkinen, Leo Lahti, Joona Lehtomaki
meta_description: ["Tools for accessing various open data sources in the Helsinki", "region in Finland. Current data sources include the Real Estate Department", "and the Environmental Services Authority."]
github_user: ropengov
package_version: 0.9.13
header_descripton: ["Tools for accessing various open data sources in the Helsinki", "region in Finland. Current data sources include the Real Estate Department", "and the Environmental Services Authority."]
---



```{r setup, echo=FALSE}
opts_knit$set(upload.fun = imgur_upload, base.url = NULL) # upload all images to imgur.com
# opts_chunk$set(fig.width=5, fig.height=5, cache=TRUE)
```

Helsinki  R tools
===========

This is an [rOpenGov](https://github.com/rOpenGov/helsinki) R package
providing tools for open data in the Helsinki region in Finland.

## Available data

The following data sets are currently available:
* [Helsinki region environmental services](#helsinki-region-environmental-services)
* [Helsinki Real Estate Department](#helsinki-real-estate-department)
* [Helsinki Service Map](#helsinki-service-map)


### Installation

Release version for general users:

```{r install, eval=FALSE}
install.packages("helsinki")
```

Development version for developers:

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("helsinki", "ropengov")
```

Load package.

```{r load}
library(helsinki)
```

Further development instructions at the [Github
page](https://github.com/rOpenGov/helsinki).

## Helsinki region environmental services

Data from Helsingin seudun ympäristöpalvelut, HSY.

### Population grid

Download population information from [HSY database](http://www.hsy.fi/seututieto/kaupunki/paikkatiedot/Sivut/Avoindata.aspx) (C) 2013 HSY, and inspect the data manually. Some rarely populated grids are censored with '99' to guarantee privacy.

```{r hsy-example, fig.width=8, fig.height=8, message=FALSE}
sp <- get_HSY_data("Vaestotietoruudukko")
df <- as.data.frame(sp)   
head(df)
```


### Helsinki building information

Information of buildings in Helsinki region. Data obtained from (C)
HSY 2013. Grid-level (500mx500m) information on building counts
(lukumaara), built area (kerrosala), usage (kayttotarkoitus), region
efficiency (aluetehokkuus).

```{r hsy-examplerak, message=FALSE}
sp <- get_HSY_data("Rakennustietoruudukko", 2013)  
df <- as.data.frame(sp)   
head(df)
```

### Helsinki building area capacity

Building area capacity per municipal region (kaupunginosittain summattu tieto rakennusmaavarannosta). Data obtained from (C) HSY 2013. 

```{r hsy-exampleramava, message=FALSE}
sp <- get_HSY_data("SeutuRAMAVA_kosa")  
df <- as.data.frame(sp)   
head(df)
```


### Further HSY data

[HSY website](http://www.hsy.fi/seututieto/kaupunki/paikkatiedot/Sivut/Avoindata.aspx) has more data that will be included in the helsinki package later.


## Helsinki Real Estate Department

Data from Helsingin kaupungin kiinteistövirasto, HKK.

Retrieve [HKK](http://kartta.hel.fi/avoindata/index.html) data sets.

### Helsinki address information

```{r hkk-addressH, message=FALSE}
dat <- get_HKK_address_data("Helsingin osoiteluettelo")
head(dat)
```

```{r hkk-addressS, message=FALSE}
dat <- get_HKK_address_data("Seudullinen osoiteluettelo")
head(dat)
```

## Helsinki Service Map

Retrieve data from [Helsinki Service Map](http://www.hel.fi/palvelukartta/?lang=en) [API](http://www.hel.fi/palvelukarttaws/rest/ver2_en.html).

```{r servicemap, message=FALSE}
# Get servicetree
pk.servicetree <- get_ServiceMap_data("servicetree")
# Get id for parks
str(pk.servicetree[[1]]$children[[7]]$children[[2]])
# Get parks data
parks.data <- get_ServiceMap_data("unit", service=25664)
# Check what data is given for the first park
str(parks.data[[1]])
```


### Licensing and Citations

This work can be freely used, modified and distributed under the
[Two-clause FreeBSD
license](http://en.wikipedia.org/wiki/BSD\_licenses). Cite Helsinki R
package and and the appropriate data provider, including a url
link. Kindly cite the R package as 'Leo Lahti, Juuso Parkkinen ja
Joona Lehtomäki (2013-2014). helsinki R package. URL:
https://github.com/rOpenGov/helsinki/'.

### Session info


This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```

