---
title: helsinki vignette
layout: tutorial_page
package_name: helsinki
package_name_show: helsinki
author: Juuso Parkkinen, Leo Lahti, Joona Lehtomaki
meta_description: ["Tools for accessing various open data sources in the Helsinki", "region in Finland. Current data sources include the Real Estate Department", "and the Environmental Services Authority."]
github_user: ropengov
package_version: 0.9.19
header_descripton: ["Tools for accessing various open data sources in the Helsinki", "region in Finland. Current data sources include the Real Estate Department", "and the Environmental Services Authority."]
---



```{r knit, eval=FALSE, echo=FALSE}
# Hot to knit this document into .md and .html
library(knitr)
opts_knit$set(base.dir = "vignettes") # Change the base dir where to save figures
knit(input="vignettes/helsinki_tutorial.Rmd", output="vignettes/helsinki_tutorial.md")
knit2html(input="vignettes/helsinki_tutorial.md", output="vignettes/helsinki_tutorial.html", options=c("use_xhtml","smartypants","mathjax","highlight_code"))
```


helsinki - tutorial
===========

This R package provides tools to access open data from the Helsinki region in Finland
as part of the [rOpenGov](http://ropengov.github.io) project.

For contact information and source code, see the [github page](https://github.com/rOpenGov/helsinki). 

## Available data sources

[Helsinki region district maps](#aluejakokartat) (Helsingin seudun aluejakokartat)
* Aluejakokartat: kunta, pien-, suur-, tilastoalueet (Helsinki region district maps)
* Äänestysaluejako: (Helsinki region election district maps)
* Source: [Helsingin kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/)

Helsinki Real Estate Department (HKK:n avointa dataa)
* Spatial data from [Helsingin kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/) availabe in the [gisfin](https://github.com/rOpenGov/gisfin) package, see [gisfin tutorial](https://github.com/rOpenGov/gisfin/blob/master/vignettes/gisfin_tutorial.md) for examples

[Helsinki region environmental services](#hsy) (HSY:n avointa dataa)
* Väestötietoruudukko (population grid)
* Rakennustietoruudukko (building information grid)
* SeutuRAMAVA (building land resource information(?))
* Source: [Helsingin seudun ympäristöpalvelut, HSY](http://www.hsy.fi/seututieto/kaupunki/paikkatiedot/Sivut/Avoindata.aspx)

[Service and event information](#servicemap)
* [Helsinki region Service Map API](http://api.hel.fi/servicemap/v1/) (Pääkaupunkiseudun Palvelukartta)
* [Helsinki Linked Event API](http://api.hel.fi/linkedevents/v0.1/)
* [Omakaupunki](http://api.omakaupunki.fi/) (requires personal API key, no examples given)

[Helsinki Region Infoshare statistics API](#hri_stats)
* [Aluesarjat (original source)](http://www.aluesarjat.fi/) (regional time series data)
* Source: [Helsinki Region Infoshare statistics API](http://dev.hel.fi/stats/)

[Economic data](#economy)
* [Taloudellisia tunnuslukuja](http://www.hri.fi/fi/data/paakaupunkiseudun-kuntien-taloudellisia-tunnuslukuja/) (economic indicators)

[Demographic data](#demography)
* [Väestöennuste 2012-2050](http://www.hri.fi/fi/data/helsingin-ja-helsingin-seudun-vaestoennuste-sukupuolen-ja-ian-mukaan-2012-2050/) (population projection 2012-2050)
 

List of potential data sources to be added to the package can be found [here](https://github.com/rOpenGov/helsinki/blob/master/vignettes/todo-datasets.md).

## Installation

Release version for general users:

```{r install, eval=FALSE}
install.packages("helsinki")
```

Development version for developers:

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("helsinki", "ropengov")
```

Load package.

```{r load, message=FALSE, warning=FALSE, results='hide'}
library(helsinki)
```

## <a name="aluejakokartat"></a>Helsinki region district maps

Helsinki region district maps (Helsingin seudun aluejakokartat) and election maps (äänestysalueet) from [Helsingin kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/) are available in the helsinki package with `data(aluejakokartat)`. The data are available as both spatial object (`sp`) and data frame (`df`). These are preprocessed in the [gisfin](https://github.com/rOpenGov/gisfin) package, and more examples can be found in the [gisfin tutorial](https://github.com/rOpenGov/gisfin/blob/master/vignettes/gisfin_tutorial.md). 

```{r aluejako_data, message=FALSE}
# Load aluejakokartat and study contents
data(aluejakokartat)
str(aluejakokartat, m=2)
```


## <a name="hsy"></a> Helsinki region environmental services

Retrieve data from [Helsingin seudun ympäristöpalvelut (HSY)](http://www.hsy.fi/seututieto/kaupunki/paikkatiedot/Sivut/Avoindata.aspx) with `get_hsy()`.

### Population grid 

Population grid (väestötietoruudukko) with 250m x 250m grid size in year 2013 contains the number of people in different age groups. The most rarely populated grids are left out (0-4 persons), and grids wiht less than 99 persons are censored with '99' to guarantee privacy.

```{r hsy_vaesto, message=FALSE, warning=FALSE}
sp.vaesto <- get_hsy(which.data="Vaestotietoruudukko", which.year=2013)
head(sp.vaesto@data)
```


### Helsinki building information

Building information grid (rakennustietoruudukko) in Helsinki region on grid-level (500m x 500m): building counts (lukumäärä), built area (kerrosala), usage (käyttötarkoitus), and region
efficiency (aluetehokkuus).

```{r hsy_rakennus, message=FALSE, warning=FALSE}
sp.rakennus <- get_hsy(which.data="Rakennustietoruudukko", which.year=2013)  
head(sp.rakennus@data)
```

### Helsinki building area capacity

Building area capacity per municipal region (kaupunginosittain summattua tietoa rakennusmaavarannosta). Plot with number of buildlings with `spplot()`.

```{r hsy_ramava, message=FALSE, warning=FALSE}
sp.ramava <- get_hsy(which.data="SeutuRAMAVA_tila", which.year=2013)  
head(sp.ramava@data)
# Values with less than five units are given as 999999999, set those to zero
sp.ramava@data[sp.ramava@data==999999999] <- 0
# Plot number of buildings for each region
spplot(sp.ramava, zcol="RAKLKM", main="Number of buildings in each 'tilastoalue'", col.regions=colorRampPalette(c('blue', 'gray80', 'red'))(100))
```


## <a name="servicemap"></a>Service and event information

Function `get_servicemap()` retrieves regional service data from the new [Service Map API](http://api.hel.fi/servicemap/v1/), that contains data from the [Service Map](http://dev.hel.fi/servicemap/).

```{r servicemap, message=FALSE, warning=FALSE}
# Search for "puisto" (park) (specify q="query")
search.puisto <- get_servicemap(query="search", q="puisto")
# Study results
str(search.puisto, m=1)
# A lot of results found (count > 1000)
# Get names for the first 20 results
sapply(search.puisto$results, function(x) x$name$fi)
# See what data is given for one service
names(search.puisto$results[[1]])
# More results could be retrieved by specifying 'page=2' etc.
```

Function `get_linkedevents()` retrieves regional event data from the new [Linked Events API](http://api.hel.fi/linkedevents/v0.1/).

```{r linkedevents, message=FALSE, warning=FALSE}
# Searh for current events
events <- get_linkedevents(query="event")
# Get names for the first 20 results
sapply(events$results, function(x) x$name$fi)
# See what data is given for the first event
names(events$results[[1]])
```


Function `get_omakaupunki()` retrieves regional service and event data from the [Omakaupunki API](http://api.omakaupunki.fi/). However, the API needs a personal key, so no examples are given here.

## <a name="hri_stats"></a> Helsinki Region Infoshare statistics API

Function `get_hri_stats()` retrieves data from the [Helsinki Region Infoshare statistics API](http://dev.hel.fi/stats/).

```{r hri_stats1, message=FALSE, warning=FALSE}
# Retrieve list of available data
stats.list <- get_hri_stats(query="")
# Show first results
head(stats.list)
```

Specify a dataset to retrieve. The output is currently a three-dimensional array.

```{r hri_stats2, message=FALSE, warning=FALSE}
# Retrieve a specific dataset
stats.res <- get_hri_stats(query=stats.list[1])
# Show the structure of the results
str(stats.res)
```

The implementation will be updated and more examples will be added in the near future.

## <a name="economy"></a> Economic data

Function `get_economic_indicators()` retrieves [economic indicator](http://www.hri.fi/fi/data/paakaupunkiseudun-kuntien-taloudellisia-tunnuslukuja/) data for Helsinki, Espoo, Vantaa and Kauniainen from years 1998-2010. 

```{r econ_ind1, message=FALSE, warning=FALSE}
# Retrieve data
ec.res <- get_economic_indicators()
# See first results
head(ec.res$data)
```


## <a name="demography"></a> Demographic data

Function `get_population_projection()` retrieves [population projection](http://www.hri.fi/fi/data/helsingin-ja-helsingin-seudun-vaestoennuste-sukupuolen-ja-ian-mukaan-2012-2050/) data for Helsinki and Helsinki region by age and gender.

```{r population_projection, message=FALSE, warning=FALSE}
# Retrieve data
pop.res <- get_population_projection()
# See first results
head(pop.res$data[1:6])
```


### Citation

**Citing the data:** See `help()` to get citation information for each data source individually.

**Citing the R package:**

```{r citation, comment=NA}
citation("helsinki")
```

### Session info


This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```

