---
title: sorvi vignette
layout: tutorial_page
package_name: sorvi
package_name_show: sorvi
author: Leo Lahti, Juuso Parkkinen, Joona Lehtomaki, Juuso Haapanen, Jussi, Paananen, Einari Happonen
meta_description: Algorithms for Finnish Open Government Data
github_user: ropengov
package_version: 0.7.12
header_descripton: Algorithms for Finnish Open Government Data
---



Finnish open government data toolkit for R
===========

This R package provides miscellaneous tools for Finnish open
government data to complement other
[rOpenGov](http://ropengov.github.io/projects) packages with a more
specific scope. We also maintain a [todo list of further data
sources](https://github.com/rOpenGov/sorvi/blob/master/vignettes/todo-datasets.md)
to be added; your
[contributions](http://louhos.github.com/contact.html) and [bug
reports and other feedback](https://github.com/ropengov/sorvi) are
welcome! For further information, see the [home
page](http://louhos.github.com/sorvi).



## Available data sources and tools

[Installation](#installation) (Asennus)  

[Finnish provinces](#provinces) (Maakuntatason informaatio)  
* [Basic province information](#provinceinfo) (Area, Population, Population Density)
* [Finnish-English province name translations](#provincetranslations)  

[Finnish municipalities](#municipality) (Kuntatason informaatio)
* [Land Survey Finland](#mml) (Maanmittauslaitos / MML)

[ID conversion tools](#conversions)
* [Municipality-Postal code conversions](#postalcodes) (Kunnat vs. postinumerot)  
* [Municipality name-ID conversions](#municipalityconversions) (Kunnat vs. kuntakoodit)
* [Municipality-province conversions](#municipality2province) (Kunnat vs. maakunnat)

[Finnish personal identification number (HETU)](#hetu) (Henkilotunnuksen kasittely)  

[Visualization tools](#visualization) (Visualisointirutiineja)



## <a name="installation"></a>Installation

We assume you have installed [R](http://www.r-project.org/). If you
use [RStudio](http://www.rstudio.com/ide/download/desktop), change the
default encoding to UTF-8. Linux users should also install
[CURL](http://curl.haxx.se/download.html).

Install the stable release version in R:

```{r install, eval=FALSE}
install.packages("sorvi")
```

Test the installation by loading the library:

```{r test, message=FALSE, warning=FALSE, eval=TRUE}
library(sorvi)
```

We also recommend setting the UTF-8 encoding:

```{r locale, eval=FALSE}
Sys.setlocale(locale="UTF-8") 
```

Brief examples of the package tools are provided below. Further
examples are available in [Louhos-blog](http://louhos.wordpress.com)
and in our [Rmarkdown blog](http://louhos.github.io/archive.html).


## <a name="provinces"></a>Province information (Maakunnat)


### <a name="provinceinfo"></a>Basic data

Source: [Wikipedia](http://fi.wikipedia.org/wiki/V%C3%A4est%C3%B6tiheys)

```{r municipality, warning=FALSE, message=FALSE, eval=TRUE}
tab <- get_province_info_wikipedia()
head(tab)
```

### <a name="provincetranslations"></a>Finnish-English translations

**Finnish-English translations for province names** (we have not been able
to solve all encoding problems yet; solutions welcome!):

```{r translate, message=FALSE, eval=TRUE}
translations <- load_sorvi_data("translations")
head(as.matrix(translations))
```


## <a name="municipality"></a>Municipality information

Finnish municipality information is available through Statistics
Finland (Tilastokeskus; see
[stafi](https://github.com/ropengov/statfi) package) and Land Survey
Finland (Maanmittauslaitos). The row names for each data set are
harmonized and can be used to match data sets from different sources,
as different data sets may carry different versions of certain
municipality names.

### <a name="mml"></a>Land Survey Finland (municipality information)

Source: [Maanmittauslaitos, MML](http://www.maanmittauslaitos.fi/aineistot-palvelut/latauspalvelut/avoimien-aineistojen-tiedostopalvelu). 

```{r municipalityMML, message=FALSE, warning=FALSE, eval=TRUE}
municipality.info.mml <- get_municipality_info_mml()
print(municipality.info.mml[1:2,])
```


## <a name="conversions"></a>Conversions

### <a name="postalcodes"></a>Postal codes vs. municipalities

Source: [Wikipedia](http://fi.wikipedia.org/wiki/Luettelo_Suomen_postinumeroista_kunnittain). The municipality names are provided also in plain ascii without special characters:

```{r postalcodes, message=FALSE, warning=FALSE, eval=TRUE}
postal.code.table <- get_postal_code_info() 
head(postal.code.table)
```


### <a name="municipality2province"></a>Municipality-Province mapping

**Map all municipalities to correponding provinces**

```{r province2, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
m2p <- municipality_to_province() 
head(m2p) # Just show the first ones
```  

**Map selected municipalities to correponding provinces:**

```{r province6, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
municipality_to_province(c("Helsinki", "Tampere", "Turku")) 
```  

**Speed up conversion with predefined info table:**

```{r province7, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
m2p <- municipality_to_province(c("Helsinki", "Tampere", "Turku"), municipality.info.mml)
head(m2p)
```  


### <a name="municipalityconversions"></a>Municipality name-ID conversion

**Municipality name to code**

```{r province3, message=FALSE, echo=TRUE, eval=TRUE}
convert_municipality_codes(municipalities = c("Turku", "Tampere"))
```  

**Municipality codes to names**

```{r province4, message=FALSE, echo=TRUE, eval=TRUE}
convert_municipality_codes(ids = c(853, 837))
```  

**Complete conversion table**

```{r province5, message=FALSE, echo=TRUE, eval=TRUE}
municipality_ids <- convert_municipality_codes()
head(municipality_ids) # just show the first entries
```



## <a name="hetu"></a>Personal identification number (HETU)

**Extract information from a Finnish personal identification number:**

```{r hetu, message=FALSE}
library(sorvi)
hetu("111111-111C")
```

**Validate Finnish personal identification number:**

```{r hetu2, fig.message=FALSE}
valid_hetu("010101-0101") # TRUE/FALSE
```



## <a name="visualization"></a>Visualization tools

Line fit with confidence smoothers (if any of the required libraries
are missing, install them with the install.packages command in R):

```{r regressionline, message=FALSE, eval=TRUE, fig.width=10, fig.height=5}
library(sorvi) 
library(plyr)
library(RColorBrewer)
library(ggplot2)
data(iris)
p <- regression_plot(Sepal.Length ~ Sepal.Width, iris) 
print(p)
```




## Licensing and Citations

This work can be freely used, modified and distributed under the 
[Two-clause BSD license](http://en.wikipedia.org/wiki/BSD\_licenses).

```{r citation, message=FALSE, eval=TRUE}
citation("sorvi")
```

## Session info

This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```




