---
title: sotkanet vignette
layout: package_page
package_name: sotkanet
package_name_show: sotkanet
author: Leo Lahti
meta_description: Sotkanet API R tools
github_user: ropengov
pacakge_version: 0.9.01
header_descripton: Sotkanet API R tools
---

```{r, eval=TRUE, echo=FALSE}
opts_chunk$set(fig.path="../_assets/img/vignettes/alm/")
knit_hooks$set(render_jekyll(highlight = "pygments"))
```

This is the [sotkanet](http://ropengov.github.com/sotkanet) R package to access data from the [Sotkanet portal](http://uusi.sotkanet.fi/portal/page/portal/etusivu/hakusivu) that provides over 2000 demographic indicators across Finland and Europe, maintained by the National Institute for Health and Welfare (THL). For more information, see [Sotkanet indicator database](http://uusi.sotkanet.fi/portal/page/portal/etusivu/tietoa_palvelusta) and [API description](http://uusi.sotkanet.fi/portal/pls/portal/!PORTAL.wwpob_page.show?_docname=22001.PDF). This package is part of [rOpenGov](http://ropengov.github.com/).


### Installation

Release version for general users:

```{r install, eval=FALSE}	
install.packages("sotkanet")
library(sotkanet)
```

Development version for developers:

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("sotkanet", "ropengov")
library(sotkanet)
```

### Listing available indicators

List available Sotkanet indicators:

```{r sotkanetIndicators, warning=FALSE, message=FALSE}
library(sotkanet) 
sotkanet.indicators <- SotkanetIndicators(type = "table")
head(sotkanet.indicators)
```

List geographical regions with available indicators:

```{r sotkanetRegions, warning=FALSE, message=FALSE}
sotkanet.regions <- SotkanetRegions(type = "table")
head(sotkanet.regions)
```


### Querying SOTKAnet indicators

Get the [indicator 10013](http://uusi.sotkanet.fi/portal/page/portal/etusivu/hakusivu/metadata?type=I&indicator=10013) from Finland (Suomi) for 1990-2012 (Eurostat employment statistics youth unemployment), and plot a graph:

```{r sotkanetData, warning=FALSE, message=FALSE}
# Get indicator data
dat <- GetDataSotkanet(indicators = 10013, years = 1990:2012, 
       		       genders = c('female', 'male', 'total'), 
		       region.category = "EUROOPPA", 
		       region = "Suomi")

# Investigate the first lines in the data
head(dat)

# Pick indicator name
indicator.name <- as.character(unique(dat$indicator.title.fi))
indicator.source <- as.character(unique(dat$indicator.organization.title.fi))

# Visualize
library(ggplot2, quietly=TRUE)
theme_set(theme_bw(20)); 
p <- ggplot(dat, aes(x = year, y = primary.value, group = gender, color = gender)) 
p <- p + geom_line() + ggtitle(paste(indicator.name, indicator.source, sep = " / ")) 
p <- p + xlab("Year") + ylab("Value") 
p <- p + theme(title = element_text(size = 10))
p <- p + theme(axis.title.x = element_text(size = 20))
p <- p + theme(axis.title.y = element_text(size = 20))
p <- p + theme(legend.title = element_text(size = 15))
print(p)
```


### Effect of municipality size

Smaller municipalities have more random variation.

```{r sotkanetVisu3, warning=FALSE, message=FALSE}
selected.indicators <- c("Väestö 31.12.", "Kuntien välinen nettomuutto / 1 000 asukasta"); 
selected.inds <- sotkanet.indicators$indicator[match(selected.indicators, sotkanet.indicators$indicator.title.fi)]
dat <- GetDataSotkanet(indicators = selected.inds, years = 2011, genders = c('total'))
datf <- dat[, c("region.title.fi", "indicator.title.fi", "primary.value")]
dw <- reshape(datf, idvar = "region.title.fi", timevar = "indicator.title.fi", direction = "wide")
names(dw) <- c("Municipality", "Population", "Migration")
p <- ggplot(dw, aes(x = log10(Population), y = Migration)) + geom_point(size = 3)
p <- p + ggtitle("Migration vs. population size") 
p <- p + theme(title = element_text(size = 15))
p <- p + theme(axis.title.x = element_text(size = 20))
p <- p + theme(axis.title.y = element_text(size = 20))
p <- p + theme(legend.title = element_text(size = 15))
print(p)
```



### Fetch all SOTKAnet indicators

This takes for a long time and is not recommended for regular
use. Save the data on your local dissk for further work.


```{r sotkanetDataAll, warning=FALSE, message=FALSE, eval=FALSE}
# These indicators have problems with R routines:
probematic.indicators <- c(1575, 1743, 1826, 1861, 1882, 1924, 1952, 2000, 2001, 2033, 2050, 3386, 3443)

# Get data for all indicators
datlist <- list()
for (ind in setdiff(sotkanet.indicators$indicator, probematic.indicators)) {
  datlist[[as.character(ind)]] <- GetDataSotkanet(indicators = ind, years = 1990:2013, genders = c('female', 'male', 'total'))
}

# Combine tables (this may require considerable time and memory 
# for the full data set)
dat <- do.call("rbind", datlist)
```

For further usage examples, see
[Louhos-blog](http://louhos.wordpress.com), and
[takomo](https://github.com/louhos/takomo/tree/master/Sotkanet).


# Licensing and Citations

### SOTKAnet data

Cite SOTKAnet and link to [http://www.sotkanet.fi](http://www.sotkanet.fi/). Also mention indicator provider. 

 * [Full license and terms of use](http://uusi.sotkanet.fi/portal/page/portal/etusivu/tietoa_palvelusta). 

Central points:

 * SOTKAnet REST API is meant for non-regular data queries. Avoid regular and repeated downloads.
 * SOTKAnet API can be used as the basis for other systems
 * Metadata for regions and indicators are under [CC-BY 3.0](http://creativecommons.org/licenses/by/3.0/) 
 * THL indicators are under [CC-BY 3.0](http://creativecommons.org/licenses/by/3.0/) 
 * Indicators provided by third parties can be used only by separate agreement!


### SOTKAnet R package

This work can be freely used, modified and distributed under the
[Two-clause FreeBSD
license](http://en.wikipedia.org/wiki/BSD\_licenses). Kindly cite the
R package as 'Leo Lahti, Einari Happonen, Juuso Parkkinen ja Joona
Lehtomäki (2013). sotkanet R package. URL:
http://ropengov.github.io/sotkanet'.


### Session info


This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```
