---
title: statfi vignette
layout: tutorial_page
package_name: statfi
package_name_show: statfi
author: Leo Lahti, Juuso Parkkinen, Joona Lehtomaki
meta_description: Statistics Finland (Tilastokeskus) Open Data R Tools
github_user: ropengov
package_version: 0.9.83
header_descripton: Statistics Finland (Tilastokeskus) Open Data R Tools
---



Statistics Finland (Tilastokeskus) R tools
===========

This R package provides tools to access open data from [Statistics
Finland](http://www.stat.fi/tup/tilastotietokannat/index_fi.html),
including about 3000 data sets from [Statistics
Finland](http://www.stat.fi/org/lainsaadanto/avoin_data.html). 

This R package is part of the [rOpenGov](http://ropengov.github.io)
project.


## Installation

Release version for general use:

```{r install, eval=FALSE}
install.packages("statfi")
library(statfi)
```

Development version (potentially unstable):

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("statfi", "ropengov")
library(statfi)
```


## Available data sets

The listings of [Statistics Finland (StatFi) open
data](http://www.stat.fi/org/lainsaadanto/avoin_data.html) are
available for browsing in PCAxis, CSV and XML format. These include
the following data collections:

 * StatFi [PCAxis](http://pxweb2.stat.fi/database/StatFin/databasetree_fi.asp) [CSV](http://pxweb2.stat.fi/database/StatFin/StatFin_rap_csv.csv) [XML](http://pxweb2.stat.fi/database/StatFin/StatFin_rap_xml.csv)  
 * Eurostat [PCAxis](http://pxweb2.stat.fi/Database/Eurostat/databasetree_fi.asp) [CSV](http://pxweb2.stat.fi/database/StatFin/StatFin_rap.csv)  
 * International statistics [PC Axis](http://pxweb2.stat.fi/Database/Kansainvalisen_tiedon_tietokanta/databasetree_fi.asp)

In summary, browse these listings to find the URL for your data set of
interest. Then use the get_statfi function to download the data in
R. For examples, see below.

### Listing the data sets in R

Download statfi open data listings in R as follows:

```{r tilastokeskus-1, warning=FALSE, message=FALSE}
# Load the library
library(statfi)

# Statistics Finland open data listing
datasets.statfi <- list_statfi_files()

# Descriptions of the first entries
head(datasets.statfi$DESCRIPTION)

# Investigate the first entry in StatFi data
print(datasets.statfi[1,])
```

This provides the list of statfi data sets. For other international
open statistics available via Statfi, [browse the data sets
manually](http://pxweb2.stat.fi/Database/Kansainvalisen_tiedon_tietokanta/databasetree_fi.asp)
to find the URL for your dataset of interest. For Eurostat data, we
recommend the [eurostat](http://github.com/ropengov/eurostat) R
package.


## Retrieving the data

Retrieve data from Statfi by defining URL of the data set. For the
listing of available data sets and their corresponding URLs, see
above.

```{r tilastokeskus-2, warning=FALSE, message=FALSE}
library(statfi)

# Define URL (see list_statfi_files() or browse manually as described above)
url <- "http://pxweb2.stat.fi/Database/StatFin/tul/tvt/2009/120_tvt_2009_2011-02-18_tau_112_fi.px"

# Download the data
df <- get_statfi(url)
df[1:3,]
```

### <a name="statfi"></a>Statistics Finland (municipality information)

Source: [Tilastokeskus](http://pxweb2.stat.fi/Database/Kuntien%20perustiedot/Kuntien%20perustiedot/Kuntaportaali.px)

```{r municipalityStatFi, message=FALSE, warning=FALSE, eval=TRUE}
# Download Statfi municipality data
municipality.info.statfi <- get_municipality_info_statfi()

# List available information fields for municipalities
names(municipality.info.statfi)
```



## Licensing and Citations

### Citing the Data

Regarding the data, kindly cite [Statfi](http://www.statfi.fi/). We
are grateful to Statistics Finland open data personnell for their
support during the R package development.

### Citing the R package

This work can be freely used, modified and distributed under the
[Two-clause FreeBSD
license](http://en.wikipedia.org/wiki/BSD\_licenses). Kindly cite the
R package as 'Leo Lahti, Juuso Parkkinen ja Joona Lehtomäki (C)
2010-2014. statfi R package. URL: http://ropengov.github.io/statfi'.


### Session info

This tutorial was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```
